Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-06-21 Manuscript/00-main.Rmd
1. Revise introduction section 2. Add some simulation results 3. Add coauthors 4. Bibfiles

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
# - num: b
  # org: Department of Applied Things, University B, City B, Country B
corres: "*Corresponding author name, This is sample corresponding address. \\email{xxx@uab}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo and Yi
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "Spike-and-slab priors have been extensively studied for variable selection, and recently been extended to function selection in generalized additive models (GAMs). Most of the previous works on fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC) algorithms to identify the promising subsets of functions. The computational burden of MCMC makes these models less feasible, especially in high-dimensional setting. Besides, most of the current spike-and-slab GAM consider an “all-in-all-out” approach for function selection, rendering less flexibility to answer if non-linear effects are necessary in the model. We propose Bayesian hierarchical generalized additive models for function selection and prediction. The proposed models employ spike-and-slab priors, mixture normal and mixture double exponential, for smooth and sparse solution. Two EM based algorithms, EM-IWLS and EM-Coordinate Descent, are developed for accelerated computation. Simulation studies and real data analysis demonstrated improved performance against the state-of-art models, mgcv and COSSO. The software implementation of the proposed methods is freely available via the R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
---

>>>>

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: D. Leann Long
  num: a
- name: Byron C. Jaeger
  num: a 
- name: AKM Fazlur Rahman
  num: a   
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
# - num: b
  # org: Department of Applied Things, University B, City B, Country B
corres: "*Corresponding author name, This is sample corresponding address. \\email{xxx@uab}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo and Yi
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "Spike-and-slab priors have been extensively studied for variable selection, and recently been extended to function selection in generalized additive models (GAMs). Most of the previous works on fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC) algorithms to identify the promising subsets of functions. The computational burden of MCMC makes these models less feasible, especially in high-dimensional setting. Besides, most of the current spike-and-slab GAM consider an “all-in-all-out” approach for function selection, rendering less flexibility to answer if non-linear effects are necessary in the model. We propose Bayesian hierarchical generalized additive models for function selection and prediction. The proposed models employ spike-and-slab priors, mixture normal and mixture double exponential, for smooth and sparse solution. Two EM based algorithms, EM-IWLS and EM-Coordinate Descent, are developed for accelerated computation. Simulation studies and real data analysis demonstrated improved performance against the state-of-art models, mgcv and COSSO. The software implementation of the proposed methods is freely available via the R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
---

### 2021-06-21 Manuscript/01-Intro.Rmd
Improved the manuscript
1. Revise introduction section 2. Add some simulation results 3. Add coauthors 4. Bibfiles

* What is the problem
  * what is the GAM
  * expanding GAM in to high-dimensional frame work
  * Why this is important
  * Why this is hard
* Previous solutions
  * What are the methods
  * What are their problems

* Introduce the wiggliness matrix

* Our method
* The implication of our method
* Fast
* Structure of the rest of the paper

Nonlinear effect modeling has long history. Early work of GAMs can date back to 1979 when @Cleveland1979 proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models has been proposed, see xxx^[TODO: insert an review citation]. With recent advancement of computing technology, many machine learning algorithms that is computationally intensive start to gainmuch popularity in outcome prediction when non-linear effects assumed. Such models generally have improved prediction accuracy over classic statistical models, but have a fatal flaw, lack of transparency and interpretability. The primary objective of biomedical studies is to understand the etiology of diseases, which requires not only the knowledge if a biomarker is predictive of the disease, but also how the biomarker cause the disease. "Black box" models could less likely to be helpful.

As a balance between interpretability and model flexibility, semi-parametric and non-parametric regression models are studied. Among those, generalized additive model (GAM) proposed in the seminal work of @Hastie1987 gains the most popularity thanks to its flexibility and easiness to incorporate many different smoothing functions. A response variable, $Y$, that is assumed to follow some exponential family distribution $EF(\mu. \phi)$, can be expressed as the summation of smooth functions of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),\qquad E[f_j(x_j)] = 0.
$$
where $g^{-1}$ is the inverse function of a smooth monotonic link function. GAM solves the model fitting problem for many distributional outcomes, such as Poisson, binomial, gamma and normal models. In the same article, @@Hastie1987 proposed the backfitting algorithm that estimates each smooth function of an additive model iteratively using partial residuals. The backfitting algorithm is compatible with almost any smoothing or modeling technique to be used as the nonparametric components. However, it is unclear how to estimate the degree of smoothness when necessary, for example for smoothing splines. The computation cost for cross-validating the smooth parameters can be expensive. @Wood2017[PP.318-320] provides more discussion on the problem.

In many area of modern biomedical research, dealing with high-dimensional data, also known as $p >> n$ problem, is common,  for example sequencing data analysis, electric health record data analysis. Many statistical models have been proposed for such problem, for example, penalized models and Bayesian hierarchical models. However, up to this point, most efforts of methodology development focus on modeling linear effect in high dimensional setting. It is still not clear what is the best practice on how to model non-linear effects in high-dimensional setting.

>>>>

Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when @Cleveland1979 proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed, see @Wood2020. With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models. However, these models have a fatal flaw, the lack of transparency and interpretability, and hence are referred to as "black box" models. Given that the primary objective of biomedical studies is to understand the etiology of diseases, "Black box" models are less helpful as they can hardly explain how the change of biomarkers is associated with the disease.

Semi-parametric regression models are a more appropriate alternative to the "black-box" models, thanks to their balancing between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of @Hastie1987, grew to be one of the most utilized modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu. \phi)$, can be expressed as the summation of smooth functions, $f_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
$$
where $g^{-1}$ is the inverse function of a smooth monotonic link function. The smooth functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classic GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

Many applications of modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There exists an extensive literature on high-dimensional linear models, via penalized models or Bayesian hierarchical models, see @Mallick2013. [TODO a sentence of sparsity. What it means in a penalized model and Bayesian model] Directly apply to additive models, spline models specifically. Problem, oversmoothing of the spline function. An more appropriate treatment would be incorporating the smooth penalty of spline functions.[TODO: a sentence about smoothness penalty]

What is the smooth penalty of spline function?

Many statistical models have been proposed for such problem, for example, penalized models and Bayesian hierarchical models. However, up to this point, most efforts of methodology development focus on modeling linear effect in high dimensional setting. It is still not clear what is the best practice on how to model non-linear effects in high-dimensional setting.

### 2021-06-21 Manuscript/03-Simulation.Rmd
1. Revise introduction section 2. Add some simulation results 3. Add coauthors 4. Bibfiles

## Simulation Study
In this section, we compare the performance of the proposed models to 3 other models: COSSO, adaptive COSSO ,mgcv. COSSO is xxx method. Adaptive COSSO is another method, mgcv is another method. 
To demonstrate the correctness and feasibility of the proposed SS-GAM,
we follow the logistic regression simulation introduced in Bai (Work in
Progress). We first generate $n=500$ training data sample with number
of predictors $p=4, 10, 50, 200$ respectively, where the predictors $X \sim MVN_{n\times p}(0, I_{P})$. We simulate the outcome data from three distributions, gaussian, poisson, and binomial with the identity link, logit link $g(x) = \log(\frac{x}{1-x})$, and log link $g(x) = log(x)$ functions to mimic the common use of generalized model. The outcome is simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
$$
for gaussian and binomial outcome and shrinked effects

@@

>>>>

# Simulation Study
In this section, we compare the performance of the proposed models to 3 other models: COSSO, adaptive COSSO ,mgcv. COSSO is xxx method. Adaptive COSSO is another method, mgcv is another method. 
To demonstrate the correctness and feasibility of the proposed SS-GAM,
we follow the logistic regression simulation introduced in Bai (Work in
Progress). We first generate $n=500$ training data sample with number
of predictors $p=4, 10, 50, 200$ respectively, where the predictors $X \sim MVN_{n\times p}(0, I_{P})$. We simulate the outcome data from three distributions, gaussian, poisson, and binomial with the identity link, logit link $g(x) = \log(\frac{x}{1-x})$, and log link $g(x) = log(x)$ functions to mimic the common use of generalized model. The outcome is simulated via the following function
$$
\mathbb{E}(Y) = g^{-1}(5 \sin(2\pi x_1) - 4 \cos(2\pi x_2 -0.5) + 6(x_3-0.5) - 5(x_4^2 -0.3))
$$
for gaussian and binomial outcome and shrinked effects

@@

For the binomial outcome shown in Figure 1, we can see that the proposed models are consistently performed better than COSSO and adaptive COSSO. Mgcv slightly outperforms the proposed models in low dimension settings(p = 4, 10), but suffers from overfitting in medium dimension setting (p=50). In addition, in the high-dimension setting (p=200), mgcv could not make prediction due to the p > n problem. Among the proposed models, we see that there is modest difference between the models that shares the same indicator variable('_spline' models) and or not sharing indicator variable (modles), where the not sharing indicator variable models have slightly edge over indicator sharing model. Moreover, the mixture double exponential model and lasso model perform substaintially better in the medium and high settings. For the remaining two outcomes, the results shows similar trends, where mixture double exponential model and ss lasso model are recommended for use.

### 2021-06-21 Manuscript/SS_GAM.tex
1. Revise introduction section 2. Add some simulation results 3. Add coauthors 4. Bibfiles

\author[a]{Boyi Guo*}
\author[a]{Nengjun Yi}

@@

Nonlinear effect modeling has long history. Early work of GAMs can date
back to 1979 when \citet{Cleveland1979} proposed local smoothing for
bivariate analysis. Since then, many mathematical and statistical models
has been proposed, see xxx\footnote{TODO: insert an review citation}.
With recent advancement of computing technology, many machine learning
algorithms that is computationally intensive start to gainmuch
popularity in outcome prediction when non-linear effects assumed. Such
models generally have improved prediction accuracy over classic
statistical models, but have a fatal flaw, lack of transparency and
interpretability. The primary objective of biomedical studies is to
understand the etiology of diseases, which requires not only the
knowledge if a biomarker is predictive of the disease, but also how the
biomarker cause the disease. ``Black box'' models could less likely to
be helpful.

As a balance between interpretability and model flexibility,
semi-parametric and non-parametric regression models are studied. Among
those, generalized additive model (GAM) proposed in the seminal work of
\citet{Hastie1987} gains the most popularity thanks to its flexibility
and easiness to incorporate many different smoothing functions. A
response variable, \(Y\), that is assumed to follow some exponential
family distribution \(EF(\mu. \phi)\), can be expressed as the summation
of smooth functions of a given \(p\)-dimensional vector of covariates
\(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),\qquad E[f_j(x_j)] = 0.
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. GAM solves the model fitting problem for many distributional
outcomes, such as Poisson, binomial, gamma and normal models. In the
same article, @\citet{Hastie1987} proposed the backfitting algorithm
that estimates each smooth function of an additive model iteratively
using partial residuals. The backfitting algorithm is compatible with
almost any smoothing or modeling technique to be used as the
nonparametric components. However, it is unclear how to estimate the
degree of smoothness when necessary, for example for smoothing splines.
The computation cost for cross-validating the smooth parameters can be
expensive. \citet[PP.318-320]{Wood2017} provides more discussion on the
problem.

@@

\hypertarget{simulation-study}{%
\subsection{Simulation Study}\label{simulation-study}}

@@

>>>>

\author[a]{Boyi Guo*}
\author[a]{D. Leann Long}
\author[a]{Byron C. Jaeger}
\author[a]{AKM Fazlur Rahman}
\author[a]{Nengjun Yi}

@@

Using additive models to interpolate nonlinear effects has a long
history. Early work of additive models can date back to 1979 when
\citet{Cleveland1979} proposed local smoothing for bivariate analysis.
Since then, many mathematical and statistical models have been proposed,
see \citet{Wood2020}. With the recent advancement of computational
technology, many machine learning algorithms that are computationally
intensive start to gain much popularity in prognostic modeling and
functional selection, especially when complex signals, for example
nonlinear effects, are assumed. Such models generally have improved
prediction accuracy over classic statistical models. However, these
models have a fatal flaw, the lack of transparency and interpretability,
and hence are referred to as ``black box'' models. Given that the
primary objective of biomedical studies is to understand the etiology of
diseases, ``Black box'' models are less helpful as they can hardly
explain how the change of biomarkers is associated with the disease.

Semi-parametric regression models are a more appropriate alternative to
the ``black-box'' models, thanks to their balancing between model
flexibility and interpretability. Among those, generalized additive
models (GAMs), proposed in the seminal work of \citet{Hastie1987}, grew
to be one of the most utilized modeling tools. In a GAM, the response
variable, \(Y\), which is assumed to follow some exponential family
distribution \(EF(\mu. \phi)\), can be expressed as the summation of
smooth functions, \(f_j, j = 1, \dots, p\), of a given \(p\)-dimensional
vector of covariates \(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. The smooth functions can take many forms and are estimated
using a pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classic GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

@@

\hypertarget{simulation-study}{%
\section{Simulation Study}\label{simulation-study}}

@@

For the binomial outcome shown in Figure 1, we can see that the proposed
models are consistently performed better than COSSO and adaptive COSSO.
Mgcv slightly outperforms the proposed models in low dimension
settings(p = 4, 10), but suffers from overfitting in medium dimension
setting (p=50). In addition, in the high-dimension setting (p=200), mgcv
could not make prediction due to the p \textgreater{} n problem. Among
the proposed models, we see that there is modest difference between the
models that shares the same indicator variable('\_spline' models) and or
not sharing indicator variable (modles), where the not sharing indicator
variable models have slightly edge over indicator sharing model.
Moreover, the mixture double exponential model and lasso model perform
substaintially better in the medium and high settings. For the remaining
two outcomes, the results shows similar trends, where mixture double
exponential model and ss lasso model are recommended for use.
